Nodal Discontinuous Galerkin Method for the Euler Equations of Fluid Dynamics

### 2017-03-21 Documents/Euler/dgEulerCurvilinear.tex
Updates to documents

We let the approximation space for the DG method, $\mathbb{V}^{k}$, to be constructed from the tensor product of one-dimensional polynomials of maximal degree $k$.  
Note that functions in $\mathbb{V}^{k}$ can be discontinuous across element interfaces.  
The semi-discrete DG problem is to find $\bU_{\DG}\in\mathbb{V}^{k}$ (which approximates $\bU$) such that
\begin{align}
  &\partial_{t}\int_{\bK}\bU_{\DG}\,v\,dV
  +\sumx\int_{\pbK^{i}}\big(\,\sqrt{\gamma}\,\widehat{\vect{F}}^{i}(\bU_{\DG})\,v\big|_{\xH^{i}}-\sqrt{\gamma}\,\widehat{\vect{F}}^{i}(\bU_{\DG})\,v\big|_{\xL^{i}}\,\big)\,\tilde{\bx}^{i} \nonumber \\
  &\hspace{24pt}
  -\sumx\int_{\bK}\vect{F}^{i}(\bU_{\DG})\,\pderiv{v}{x^{i}}\,dV
  =\int_{\bK}\bG(\bU_{\DG})\,v\,dV
  \label{eq:semidiscreteDG}
\end{align}
for all $v\in\mathbb{V}^{k}$ and all $\bK\in\mathcal{T}$.

@@

@@

Inserting \eqref{eq:conservedNodalExpansion} into \eqref{eq:semidiscreteDG}, with $v(\bx)=\ell_{\bk}(\bx)$ and the quadratures defined above, we obtain
\begin{align}
  \pd{}{t}\int_{\bK}\bU_{\DG}\,v\,dV
  &\approx w_{\bk}\,\sqrt{\gamma}_{\bk}\,\pd{}{t}\bU_{\bk}\,|\bK|
\end{align}
for the time derivative term, and
\begin{align}
  \int_{\bK}\bG(\bU_{\DG})\,v\,dV
  &\approx w_{\bk}\,\sqrt{\gamma}_{\bk}\,\bG(\bU_{\bk})\,|\bK|.
\end{align}
for the source term.  
The gravitational source terms contain the derivative to the potential, which warrants some additional exposition.  
We choose a local representation of the potential (obtained by evaluating the potential in the nodes in $\bK$ using the FEM solution); i.e.,
\begin{equation}
  \Phi(\bx)=\sum_{\bi=\vect{1}}^{\bN}\Phi_{\bi}\,\ell_{\bi}(\bx).  
\end{equation}
Differentiating with respect to $x^{j}$ gives
\begin{equation}
  \pderiv{\Phi}{x^{j}}\Big|_{\bx_{\bk}}
  =\f{1}{|K^{j}|}\sum_{\bi=\vect{1}}^{\bN}\pderiv{\ell_{\bi}}{\eta^{j}}(\eta_{\bk})\,\Phi_{\bi}
  =\f{1}{|K^{j}|}\sum_{i_{j}=1}^{N}\pderiv{\ell_{i_{j}}}{\eta^{j}}(\eta_{k_{j}}^{j})\,\Phi(x_{i_{j}}^{j},\tilde{\bx}_{\tilde{\bk}_{j}}^{j}).  
\end{equation}
This expression is used to evaluate the gravitational sources in the momentum and energy equations.

For the surface integrals, we obtain, e.g., 
\begin{align}
  %
  \int_{\tilde{\bK}^{1}}v\,\widehat{\bF}^{1}|_{x_{\Hgh}^{1}}\,dA^{1}
  &\approx w_{\tilde{\bk}_{1}}\,\ell_{k_{1}}(+1)\,\widehat{\bF}^{1}(x_{\Hgh}^{1},\tilde{x}_{\tilde{\bk}_{1}}^{1})\,
  \tau^{1}(x_{\Hgh}^{1},\tilde{x}_{\tilde{\bk}_{1}}^{1})\,|\tilde{\bK}^{1}| \nonumber \\
  &\hspace{12pt}
  =w_{\tilde{\bk}_{1}}\,\ell_{k_{1}}(+1)\,\widehat{\bF}^{1}(x_{\Hgh}^{1},\tilde{x}_{\tilde{\bk}_{1}}^{1})\,
  a(x_{\Hgh}^{1})\,b(x_{\Hgh}^{1})\,c(x_{k_{2}}^{2})\,\dx^{2}\,\dx^{3}, \\
  %
  \int_{\tilde{\bK}^{2}}v\,\widehat{\bF}^{2}|_{x_{\Hgh}^{2}}\,dA^{2}
  &\approx w_{\tilde{\bk}_{2}}\,\ell_{k_{2}}(+1)\,\widehat{\bF}^{2}(x_{\Hgh}^{2},\tilde{x}_{\tilde{\bk}_{2}}^{2})\,
  \tau^{2}(x_{\Hgh}^{2},\tilde{x}_{\tilde{\bk}_{2}}^{2})\,|\tilde{\bK}^{2}| \nonumber \\
  &\hspace{12pt}
  =w_{\tilde{\bk}_{2}}\,\ell_{k_{2}}(+1)\,\widehat{\bF}^{2}(x_{\Hgh}^{2},\tilde{x}_{\tilde{\bk}_{2}}^{2})\,
  b(x_{k_{1}}^{1})\,c(x_{\Hgh}^{2})\,\dx^{1}\,\dx^{3}, \\
  %
  \int_{\tilde{\bK}^{3}}v\,\widehat{\bF}^{3}|_{x_{\Hgh}^{3}}\,dA^{3}
  &\approx w_{\tilde{\bk}_{3}}\,\ell_{k_{3}}(+1)\,\widehat{\bF}^{3}(x_{\Hgh}^{3},\tilde{x}_{\tilde{\bk}_{3}}^{3})\,
  \tau^{3}(x_{\Hgh}^{3},\tilde{x}_{\tilde{\bk}_{3}}^{3})\,|\tilde{\bK}^{3}| \nonumber \\
  &\hspace{12pt}
  =w_{\tilde{\bk}_{3}}\,\ell_{k_{3}}(+1)\,\widehat{\bF}^{3}(x_{\Hgh}^{3},\tilde{x}_{\tilde{\bk}_{3}}^{3})\,
  a(x_{k_{1}}^{1})\,\dx^{1}\,\dx^{2}.  
\end{align}
Finally, the `volume terms' become
\begin{align}
  %
  \int_{\bK}\bF^{1}\,\pderiv{v}{x^{1}}\,dV
  &\approx w_{\tilde{\bk}_{1}}\sum_{q_{1}=1}^{N}w_{q_{1}}\,\bF^{1}(x_{q_{1}}^{1},\tilde{\bx}_{\tilde{\bk}_{1}}^{1})\,
  \pderiv{\ell_{k_{1}}}{\eta^{1}}(\eta_{q_{1}}^{1})\,\sqrt{\gamma}(x_{q_{1}}^{1},\tilde{\bx}_{\tilde{\bk}_{1}}^{1})\,|\tilde{\bK}^{1}| \nonumber \\
  &\hspace{12pt}
  =w_{\tilde{\bk}_{1}}\,c(x_{k_{2}}^{2})\,\dx^{2}\,\dx^{3}
  \sum_{q_{1}=1}^{N}w_{q_{1}}\,\bF^{1}(x_{q_{1}}^{1},\tilde{\bx}_{\tilde{\bk}_{1}}^{1})\,
  \pderiv{\ell_{k_{1}}}{\eta^{1}}(\eta_{q_{1}}^{1})\,a(x_{q_{1}}^{1})\,b(x_{q_{1}}^{1}), \\
  %
  \int_{\bK}\bF^{2}\,\f{1}{a}\pderiv{v}{x^{2}}\,dV
  &\approx w_{\tilde{\bk}_{2}}\sum_{q_{2}=1}^{N}w_{q_{2}}\,\bF^{2}(x_{q_{2}}^{2},\tilde{\bx}_{\tilde{\bk}_{2}}^{2})\,
  \f{1}{a(x_{k_{1}}^{1})}\pderiv{\ell_{k_{2}}}{\eta^{2}}(\eta_{q_{2}}^{2})\,\sqrt{\gamma}(x_{q_{2}}^{2},\tilde{\bx}_{\tilde{\bk}_{2}}^{2})\,|\tilde{\bK}^{2}| \\
  &\hspace{12pt}
  =w_{\tilde{\bk}_{2}}\,b(x_{k_{1}}^{1})\,\dx^{1}\,\dx^{3}\sum_{q_{2}=1}^{N}w_{q_{2}}\,\bF^{2}(x_{q_{2}}^{2},\tilde{\bx}_{\tilde{\bk}_{2}}^{2})\,
  \pderiv{\ell_{k_{2}}}{\eta^{2}}(\eta_{q_{2}}^{2})\,c(x_{q_{2}}^{2}), \\
  %
  \int_{\bK}\bF^{3}\,\f{1}{bc}\pderiv{v}{x^{3}}\,dV
  &\approx w_{\tilde{\bk}_{3}}\sum_{q_{3}=1}^{N}w_{q_{3}}\,\bF^{3}(x_{q_{3}}^{3},\tilde{\bx}_{\tilde{\bk}_{3}}^{3})\,
  \f{1}{b(x_{k_{1}}^{1})\,c(x_{k_{2}}^{2})}\pderiv{\ell_{k_{3}}}{\eta^{3}}(\eta_{q_{3}}^{3})\,\sqrt{\gamma}(x_{q_{3}}^{3},\tilde{\bx}_{\tilde{\bk}_{3}}^{3})\,|\tilde{\bK}^{3}| \nonumber \\
  &\hspace{12pt}
  =w_{\tilde{\bk}_{3}}\,a(x_{k_{1}}^{1})\,\dx^{1}\,\dx^{2}\sum_{q_{3}=1}^{N}w_{q_{3}}\,\bF^{3}(x_{q_{3}}^{3},\tilde{\bx}_{\tilde{\bk}_{3}}^{3})\,
  \pderiv{\ell_{k_{3}}}{\eta^{3}}(\eta_{q_{3}}^{3}).  
\end{align}

Combining the terms above with forward Euler time integration, we obtain
\begin{align}
  \bU_{\bk}^{n+1}
  &=\bU_{\bk}^{n} \nonumber \\
  &\hspace{12pt}
  -\f{\dt}{w_{k_{1}}\,a(x_{k_{1}}^{1})\,b(x_{k_{1}}^{1})\,\dx^{1}}
  \Big(\,
    a(x_{\Hgh}^{1})\,b(x_{\Hgh}^{1})\,\widehat{\bF}^{1,n}(x_{\Hgh}^{1},\tilde{x}_{\tilde{\bk}_{1}}^{1})\,\ell_{k_{1}}(+1) \nonumber \\
    &\hspace{128pt}
    -a(x_{\Low}^{1})\,b(x_{\Low}^{1})\,\widehat{\bF}^{1,n}(x_{\Low}^{1},\tilde{x}_{\tilde{\bk}_{1}}^{1})\,\ell_{k_{1}}(-1)
  \,\Big) \nonumber \\
  &\hspace{12pt}
  +\f{\dt}{w_{k_{1}}\,a(x_{k_{1}}^{1})\,b(x_{k_{1}}^{1})\,\dx^{1}}
  \sum_{q_{1}=1}^{N}w_{q_{1}}\,a(x_{q_{1}}^{1})\,b(x_{q_{1}}^{1})\,\bF^{1,n}(x_{q_{1}}^{1},\tilde{\bx}_{\tilde{\bk}_{1}}^{1})\,
  \pderiv{\ell_{k_{1}}}{\eta^{1}}(\eta_{q_{1}}^{1}) \nonumber \\
  &\hspace{12pt}
  -\f{\dt}{w_{k_{2}}\,a(x_{k_{1}}^{1})\,c(x_{k_{2}}^{2})\,\dx^{2}}
  \Big(\,
    c(x_{\Hgh}^{2})\,\widehat{\bF}^{2,n}(x_{\Hgh}^{2},\tilde{x}_{\tilde{\bk}_{2}}^{2})\,\ell_{k_{2}}(+1) \nonumber \\
    &\hspace{128pt}
    -c(x_{\Low}^{2})\,\widehat{\bF}^{2,n}(x_{\Low}^{2},\tilde{x}_{\tilde{\bk}_{2}}^{2})\,\ell_{k_{2}}(-1)
  \,\Big) \nonumber \\
  &\hspace{12pt}
  +\f{\dt}{w_{k_{2}}\,a(x_{k_{1}}^{1})\,c(x_{k_{2}}^{2})\,\dx^{2}}
  \sum_{q_{2}=1}^{N}w_{q_{2}}\,\bF^{2}(x_{q_{2}}^{2},\tilde{\bx}_{\tilde{\bk}_{2}}^{2})\,\pderiv{\ell_{k_{2}}}{\eta^{2}}(\eta_{q_{2}}^{2})\,c(x_{q_{2}}^{2}) \nonumber \\
  &\hspace{12pt}
  -\f{\dt}{w_{k_{3}}\,b(x_{k_{1}}^{1})\,c(x_{k_{2}}^{2})\,\dx^{3}}
  \Big(\,
    \widehat{\bF}^{3,n}(x_{\Hgh}^{3},\tilde{x}_{\tilde{\bk}_{3}}^{3})\,\ell_{k_{3}}(+1)
    -\widehat{\bF}^{3,n}(x_{\Low}^{3},\tilde{x}_{\tilde{\bk}_{3}}^{3})\,\ell_{k_{3}}(-1)
  \,\Big) \nonumber \\
  &\hspace{12pt}
  +\f{\dt}{w_{k_{3}}\,b(x_{k_{1}}^{1})\,c(x_{k_{2}}^{2})\,\dx^{3}}
  \sum_{q_{3}=1}^{N}w_{q_{3}}\,\bF^{3}(x_{q_{3}}^{3},\tilde{\bx}_{\tilde{\bk}_{3}}^{3})\,
  \pderiv{\ell_{k_{3}}}{\eta^{3}}(\eta_{q_{3}}^{3})
  +\dt\,\bS(\bU_{\bk}^{n}).  
\end{align}

@@

\section{Slope Limiters}

>>>>

We let the approximation space for the DG method, $\mathbb{V}^{k}$, to be constructed from the tensor product of one-dimensional polynomials of maximal degree $k$.  
Note that functions in $\mathbb{V}^{k}$ can be discontinuous across element interfaces.  
The semi-discrete DG problem is to find $\bU_{\DG}\in\mathbb{V}^{k}$ (which approximates $\bU$) such that
\begin{align}
  &\partial_{t}\int_{\bK}\bU_{\DG}\,v\,dV
  +\sumx\int_{\pbK^{i}}\big(\,\sqrt{\gamma}\,\widehat{\vect{F}}^{i}(\bU_{\DG})\,v\big|_{\xH^{i}}-\sqrt{\gamma}\,\widehat{\vect{F}}^{i}(\bU_{\DG})\,v\big|_{\xL^{i}}\,\big)\,d\tilde{\bx}^{i} \nonumber \\
  &\hspace{24pt}
  -\sumx\int_{\bK}\vect{F}^{i}(\bU_{\DG})\,\pderiv{v}{x^{i}}\,dV
  =\int_{\bK}\bG(\bU_{\DG})\,v\,dV
  \label{eq:semidiscreteDG}
\end{align}
for all $v\in\mathbb{V}^{k}$ and all $\bK\in\mathcal{T}$.

@@

\subsubsection{Notation and Definitions}

@@

\subsubsection{Explicit Expressions}

Inserting \eqref{eq:conservedNodalExpansion} into \eqref{eq:semidiscreteDG}, with $v(\bx)=\ell_{\bk}(\bx)$ and the quadratures defined above, we obtain
\begin{align}
  \pd{}{t}\int_{\bK}\bU_{\DG}\,v\,dV
  &\approx w_{\bk}\,\sqrt{\gamma}_{\bk}\,\pd{}{t}\bU_{\bk}\,|\bK|
\end{align}
for the time derivative term, where $|\bK|=\prod_{i=1}^{d}\Delta x^{i}$.  
The integration is approximate since we use the Legendre-Gauss quadrature rule with the nodal points given by the expansion in Eq.~\eqref{eq:conservedNodalExpansion}.  
This leads to a diagonal mass matrix, and simplifies the implementation.  
Similarly, we obtain
\begin{align}
  \int_{\bK}\bG(\bU_{\DG})\,v\,dV
  &\approx w_{\bk}\,\sqrt{\gamma}_{\bk}\,\bG(\bU_{\bk})\,|\bK|.
\end{align}
for the source term.  
The gravitational source terms warrants some additional exposition, which we defer to Section~\ref{sec:sourceTerms}.

For the surface integrals, we obtain, e.g., 
\begin{equation}
  \int_{\pbK^{i}}\sqrt{\gamma}\,\widehat{\vect{F}}^{i}\,v\big|_{\xH^{i}}\,d\tilde{\bx}^{i}
  \approx w_{\tilde{\bk}_{i}}\,\sqrt{\gamma}(x_{\Hgh}^{i},\tilde{\bx}_{\tilde{\bk}_{i}}^{i})\,\widehat{\bF}^{i}(x_{\Hgh}^{i},\tilde{\bx}_{\tilde{\bk}_{i}}^{i})\,\ell_{k_{i}}(\eta_{\Hgh}^{i})\,|\tilde{\bK}^{i}|,
\end{equation}
where $|\tilde{\bK}^{i}|=\prod_{j\ne i}\Delta x^{j}$.  
(Note: we do not use Einstein's summation convention in the numerical expressions presented in this section.  
Summation over indices will be explicitly indicated.)  
Finally, the volume term becomes
\begin{equation}
  \int_{\bK}\bF^{i}\,\pderiv{v}{x^{i}}\,dV
  \approx w_{\tilde{\bk}_{i}}\sum_{q_{i}=1}^{N}
  w_{q_{i}}\,\sqrt{\gamma}(x_{q_{i}}^{i},\tilde{\bx}_{\tilde{\bk}_{i}}^{i})\,\bF^{i}(x_{q_{i}}^{i},\tilde{\bx}_{\tilde{\bk}_{i}}^{i})\,\pderiv{\ell_{k_{i}}}{\eta^{i}}(\eta_{q_{i}}^{i})\,|\tilde{\bK}^{i}|.
\end{equation}
Combining the terms and dividing though by $w_{\bk}\,\sqrt{\gamma}_{\bk}\,|\bK|$, we obtain the semi-discrete for of the Euler equations
\begin{align}
  \pd{}{t}\bU_{\bk}
  &=
  -\f{1}{\sqrt{\gamma}_{\bk}}
  \sum_{i=1}^{d}\f{1}{w_{k_{i}}\,\Delta x^{i}}\,
  \Big(\,
    \sqrt{\gamma}(x_{\Hgh}^{i},\tilde{\bx}_{\tilde{\bk}_{i}}^{i})\,\widehat{\bF}^{i}(x_{\Hgh}^{i},\tilde{\bx}_{\tilde{\bk}_{i}}^{i})\,\ell_{k_{i}}(\eta_{\Hgh}^{i})
    -\sqrt{\gamma}(x_{\Low}^{i},\tilde{\bx}_{\tilde{\bk}_{i}}^{i})\,\widehat{\bF}^{i}(x_{\Low}^{i},\tilde{\bx}_{\tilde{\bk}_{i}}^{i})\,\ell_{k_{i}}(\eta_{\Low}^{i})
  \,\Big) \nonumber \\
  &\hspace{12pt}
  +\f{1}{\sqrt{\gamma}_{\bk}}
  \sum_{i=1}^{d}\f{1}{w_{k_{i}}\,\Delta x^{i}}\sum_{q_{i}=1}^{N}w_{q_{i}}\,\sqrt{\gamma}(x_{q_{i}}^{i},\tilde{\bx}_{\tilde{\bk}_{i}}^{i})\,\bF^{i}(x_{q_{i}}^{i},\tilde{\bx}_{\tilde{\bk}_{i}}^{i})\,\pderiv{\ell_{k_{i}}}{\eta^{i}}(\eta_{q_{i}}^{i})
  +\bG(\bU_{\bk})
  \label{eq:semidiscreteDiscretized}
\end{align}
Equation~\eqref{eq:semidiscreteDiscretized} provides the basis for implementation in our code.

\subsubsection{Reconstruction to Element interfaces}

\begin{equation}
  \bU_{\DG}(\eta_{\Low}^{i},\tilde{\vect{\eta}}_{\tilde{\bq}_{i}}^{i})
  =\sum_{\bk=\vect{1}}^{\bN}\bU_{\bk}\,\ell_{\bk}(\eta_{\Low}^{i},\tilde{\vect{\eta}}_{\tilde{\bq}_{i}}^{i})
  =\sum_{k_{i}=1}^{N}\sum_{\tilde{\bk}_{i}=\vect{1}}^{\bN}\bU_{(k_{i},\tilde{\bk}_{i})}\,\ell_{k_{i}}(\eta_{\Low}^{i})\,\ell_{\tilde{\bk}_{i}}(\tilde{\vect{\eta}}_{\tilde{\bq}_{i}}^{i})
\end{equation}

\subsubsection{The One-Dimensional Case}

Here we simplify the DG discretization in Eq.~\eqref{eq:semidiscreteDiscretized} to the one-dimensional ($d=1$) case to make the notation more transparent.  
With $d=1$, we substitute $\bk\to k_{d}\to k$, $x^{d}\to x$, $q_{i}\to q$, etc., so that Eq.~\eqref{eq:semidiscreteDiscretized} reduces to
\begin{align}
  \pd{}{t}\bU_{k}
  &=-\f{1}{\sqrt{\gamma}_{k}}
  \f{1}{w_{k}\,\Delta x}\,
  \Big(\,
    \sqrt{\gamma}(x_{\Hgh})\,\widehat{\bF}(x_{\Hgh})\,\ell_{k}(\eta_{\Hgh})
    -\sqrt{\gamma}(x_{\Low})\,\widehat{\bF}(x_{\Low})\,\ell_{k}(\eta_{\Low})
  \,\Big) \nonumber \\
  & \hspace{12pt}
  +\f{1}{\sqrt{\gamma}_{k}}
  \f{1}{w_{k}\,\Delta x}\sum_{q=1}^{N}w_{q}\,\sqrt{\gamma}(x_{q})\,\bF(x_{q})\,\pderiv{\ell_{k}}{\eta}(\eta_{q})
  +\bG(\bU_{k}).
  \label{eq:semidiscreteDiscretized1D}
\end{align}

\subsection{Source Terms}
\label{sec:sourceTerms}

@@

\section{Polynomial Limiting}

\subsection{Slope Limiters}

\subsection{Positivity Limiters}

\section{Equation of State (EoS)}

\section{Gravity Solver: Finite Element Method}

To solve Poisson's equation given by Eq.~\eqref{eq:poisson}, we use the finite element method (FEM).  
\ee{Nick fills in here}.

\section{Benchmark Problems}

\subsection{Smooth Problem: Linear Waves}

\subsection{Riemann Problems}

\subsubsection{Interacting Blast Waves}

\subsubsection{Shock-Entropy Wave Interaction}

\subsubsection{Sedov Blast Wave}

\subsubsection{Riemann Problem with Tabulated Nuclear EoS}

\subsection{Gravity Problems}

\subsubsection{Homogeneous Sphere}

\subsubsection{Hydrostatic Polytrope}

\subsubsection{Evrard's Collapse Test}

\subsubsection{Stationary Accretion Shock}

\section{Miniapp: Adiabatic Collapse of Stellar Iron Core}
